Symbolic Weighted Language Models and Parsing over Infinite Alphabets

### 2021-05-05 main.tex
suppr. epsilon

\noindent 
A \emph{label theory} over $\Sigma$ and $\Delta$
is made of 4 recursively enumerable sets:
%$\Phi_0$, \Phi_\Sigma \uplus \Phi_\Delta \uplus \Phi_{\Sigma, \Delta}$ 
$\Phi_\epsilon \subseteq \Semiring$, % containing constant functions valued in $\Semiring$, 
$\Phi_\Sigma$ and $\Phi_\Delta$, 
containing unary functions in $\Sigma \to \Semiring$, resp. $\Delta \to \Semiring$, 
and $\Phi_{\Sigma, \Delta}$  containing binary functions in $\Sigma \times \Delta \to \Semiring$.
Moreover, these sets are required to be closed under the operators 
$\oplus$ and $\otimes$ of~$\Semiring$.
More precisely, for all $\phi, \phi' \in \Phi_\Sigma$ all 
$\psi, \psi' \in \Phi_\Delta$, 
and $\eta, \eta' \in \Phi_{\Sigma, \Delta}$, the function
\begin{description}
\item $\phi \otimes \phi' : x \mapsto \phi(x) \otimes \phi'(x)$ belongs to $\Phi_\Sigma$, 
\item $\psi \otimes \psi' : y \mapsto \psi(y) \otimes \psi'(y)$ belongs to $\Phi_\Delta$,
\item $\phi \otimes \eta : x, y \mapsto \phi(x) \otimes \eta(x, y)$ belongs to $\Phi_{\Sigma, \Delta}$,
\item $\eta \otimes \psi : x, y \mapsto \eta(x, y) \otimes \psi(y)$ belongs to $\Phi_{\Sigma, \Delta}$,
\item $\eta \otimes \eta' : x, y \mapsto \eta(x, y) \otimes \eta'(x, y)$ belongs to $\Phi_{\Sigma, \Delta}$.
\end{description}
The same also holds for the binary sum operator $\oplus$.

@@

\noindent
The \emph{Symbolic Weighted Automata} %$A = \< Q, \init, \weight, \final >$
over $\Sigma$ and $\Semiring$ 
are defined similarly as the transducers of Definition~\ref{def:SWT}, 
by simply omitting the output symbols.
%
\begin{definition}
A \emph{symbolic-weighted automaton} (\SWA)
over the input alphabet~$\Sigma$
and the semiring $\Semiring$ is a tuple
$A = \< Q, \init, \vec{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\vec{\wei}$ is a tuplet made of the 2 of transition functions 
$\wei_\epsilon$, and $\wei_\Sigma$ from $Q \times Q$ into respectively
$\Phi_\epsilon$, and $\Phi_\Sigma$.
\end{definition}

\noindent
The 2 above transition functions are synthesized into 
a function~$\wei$ 
from~$Q \times (\Sigma \cup \{ \epsilon \}) \times Q$ into~$\Semiring$.
%
When $\wei_\epsilon(q, q') = \zero$ for all $q, q' \in Q$, 
the automaton $A$ is called \emph{without $\epsilon$-transitions}.

@@

\begin{proposition}
Given a \SWT $T$ 
over $\Sigma$, $\Delta$ and $\Semiring$, 
and a word $s \in \Sigma^*$, 
one can construct a \SWA 
$A_{s, T}$ such that for all $t \in \Delta^*$, 
$A_{s, T}(t) = T(s, t)$.
\end{proposition}
The construction time and size for $A_{s, T}$ are $O(| s | . \| T \|)$.

\begin{proposition}
...removal of $\epsilon$ transitions for \SWA...
\end{proposition}

>>>>

\noindent 
A \emph{label theory} over $\Sigma$ and $\Delta$
is made of 4 recursively enumerable sets:
%$\Phi_0$, \Phi_\Sigma \uplus \Phi_\Delta \uplus \Phi_{\Sigma, \Delta}$ 
$\Phi_\epsilon \subseteq \Semiring$, % containing constant functions valued in $\Semiring$, 
$\Phi_\Sigma$ and $\Phi_\Delta$, 
containing unary functions in $\Sigma \to \Semiring$, resp. $\Delta \to \Semiring$, 
and $\Phi_{\Sigma, \Delta}$  containing binary functions in $\Sigma \times \Delta \to \Semiring$.
Moreover, these sets are required to be closed under the operators 
$\oplus$ and $\otimes$ of~$\Semiring$.
More precisely, for all 
$\alpha, \alpha' \in \Phi_\epsilon$,
all $\phi, \phi' \in \Phi_\Sigma$,
all $\psi, \psi' \in \Phi_\Delta$, 
and $\eta, \eta' \in \Phi_{\Sigma, \Delta}$, the function
\begin{description}
\item $\phi \otimes \alpha : x \mapsto \phi(x) \otimes \alpha$, 
      and $\alpha \otimes \phi' : x \mapsto \alpha \otimes \phi'(x)$ belong to $\Phi_\Sigma$, 
\item $\phi \otimes \phi' : x \mapsto \phi(x) \otimes \phi'(x)$ belongs to $\Phi_\Sigma$, 
\item $\psi \otimes \psi' : y \mapsto \psi(y) \otimes \psi'(y)$ belongs to $\Phi_\Delta$,
\item $\phi \otimes \eta : x, y \mapsto \phi(x) \otimes \eta(x, y)$ belongs to $\Phi_{\Sigma, \Delta}$,
\item $\eta \otimes \psi : x, y \mapsto \eta(x, y) \otimes \psi(y)$ belongs to $\Phi_{\Sigma, \Delta}$,
\item $\eta \otimes \eta' : x, y \mapsto \eta(x, y) \otimes \eta'(x, y)$ belongs to $\Phi_{\Sigma, \Delta}$.
\end{description}
The same also holds for the binary sum operator $\oplus$.

@@

\noindent
The \emph{Symbolic Weighted Automata} %$A = \< Q, \init, \weight, \final >$
over $\Sigma$ and $\Semiring$ 
are defined similarly as the transducers of Definition~\ref{def:SWT}, 
by simply omitting the output symbols.
%
\begin{definition}
A \emph{symbolic-weighted automaton} (\SWA)
over the input alphabet~$\Sigma$
and the semiring $\Semiring$ is a tuple
$A = \< Q, \init, \vec{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\vec{\wei}$ is a tuplet made of the 2 of transition functions 
$\wei_\epsilon$ and $\wei_\Sigma$ from $Q \times Q$ into respectively
$\Phi_\epsilon$ and $\Phi_\Sigma$.
\end{definition}

\noindent
The above transition functions $\wei_\epsilon$ and $\wei_\Sigma$
are synthesized into a function~$\wei$ 
from~$Q \times (\Sigma \cup \{ \epsilon \}) \times Q$ into~$\Semiring$.
%
When $\wei_\epsilon(q, q') = \zero$ for all $q, q' \in Q$, 
the automaton~$A$ is called \emph{without $\epsilon$-transitions}.

@@

\begin{proposition}
Given a \SWT $T$ 
over $\Sigma$, $\Delta$ and $\Semiring$, 
and a word $s \in \Sigma^*$, 
there exists a \SWA 
$A_{T, s}$ such that for all $t \in \Delta^*$, 
$A_{T, s}(t) = T(s, t)$.
\end{proposition}
The construction time and size for $A_{s, T}$ are $O(| s | . \| T \|)$.

% gen. to transducers
\begin{proposition}
Given a \SWA $A$ over $\Sigma$ and $\Semiring$, 
there exists a \SWA $A'$ without $\epsilon$-transitions 
such that for all $s \in \Sigma^*$, $A'(s) = A(s)$.
\end{proposition}
%
\begin{proof}     
Let $A = \< Q, \init, \<\wei_\epsilon, \wei_\Sigma>, \final >$.
The automaton $A'$ is $\< Q, \init, \<\wei'_\epsilon, \wei'_\Sigma>, \final >$,
where, for all $q, q' \in Q$,
$\wei'_\epsilon(q, q') = \zero$  
and 
% \[
% \begin{array}{rcccl}
% \wei'_\Sigma(q, q') & = &  & \displaystyle\bigoplus_{q_1 \in Q} &
%                \wei_\Sigma(q, q_1) \otimes  \wei_\epsilon(q_1, q')\\
%                     &   & \oplus  & \displaystyle\bigoplus_{q_2 \in Q} &
%                \wei_\epsilon(q, q_2) \otimes \wei_\Sigma(q_2, q')\\
% \end{array}
% \]
\[
\wei'_\Sigma(q, q') = 
\displaystyle\bigoplus_{q_1 \in Q} \wei_\Sigma(q, q_1) \otimes \wei_\epsilon(q_1, q')
\oplus  \displaystyle\bigoplus_{q_2 \in Q} \wei_\epsilon(q, q_2) \otimes \wei_\Sigma(q_2, q')
\]
In the above definition of $\wei'_\Sigma$ we use the operators 
$\oplus$ and $\otimes$ on $\Phi_\Sigma$ described in Section~\ref{section:symbols}.
\qed
\end{proof}
